Davos: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-09-25 paper/main.tex
refined first few illustrative example snippet descriptions:

Across different versions of a given package, various modules, functions,
and other objects may be updated, removed, renamed, or otherwise altered. In
addition to changing the behaviors of active computations, these changes can
render saved objects created using one version of a package incompatible with
other versions of the same package. For example, the popular
\texttt{pandas}~\cite{McKi10} library used to include the \texttt{Panel} data
structure for storing 3-dimensional arrays. Since version 0.20.0, however, the
\texttt{Panel} class has been deprecated, and in version 0.25.0, it was removed
entirely. Suppose a user had a dataset stored in a \texttt{Panel} object
(created using an older version of \texttt{pandas}) and had saved it to their
disk (e.g., for later reuse or to share with other users) by serializing the
\texttt{Panel} with Python's \texttt{pickle} protocol. The \texttt{pickle}
protocol is a popular built-in method of persisting data in Python that allows
users to save, share, and load arbitrary objects. However, in order to
successfully ``unpickle'' (i.e., load and restore) a ``pickled'' (i.e., previously saved)
object, that object's class must be defined in and importable from the same
module as it was when the object was originally saved. Thus, because of the \texttt{Panel} class's
removal, the user's dataset could not be read by any version of \texttt{pandas}
from 0.25.0 onward. These incompatibilities are also not limited solely to
traditional forms of data. For example, saved model states and other objects
may reference modules, functions, attributes, classes, or other objects that
may not be identical (or even present) across all versions of their associated
packages.

The example provided in Figure~\ref{fig:illustrative-example} demonstrates how
Davos can be used to circumvent these incompatibilities by
temporarily switching between different versions of the same package within a single runtime.
%carefully controlling which versions of each package are used in different parts of the notebook.
The example shows how a dataset and model that require
now-incompatible components of the \texttt{pandas} and
\texttt{scikit-learn}~\cite{PedrEtal11} libraries may be loaded in (using older
versions of each package) and used alongside more recent versions of each
package that provide new and improved functionality. When included at the top
of an IPython notebook, the code in Figure~\ref{fig:illustrative-example}
ensures that these objects will be loaded successfully and analyzed using the
same set of package versions, no matter when or by whom the notebook is run.

After installing and importing Davos (lines 1--2), we first use the \texttt{davos.require\_python()} function to constrain the Python version used to run the code below (see Sec.~\ref{subsec:toplevel}).
As described above, the example code in Figure \ref{fig:illustrative-example} uses two different versions of the \texttt{pandas} library: first, an older one needed to access a dataset saved in an outmoded format, then a newer one throughout the remainder of the notebook.
Line 6 therefore ensures upfront that the notebook's Python version falls within the overlap between the ranges of Python versions that these two versions of \texttt{pandas} support. If it does not, the function in line 6 will raise an error that includes a message to this effect (lines 4--5).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example1}
\end{center}
Next, in lines 8--9 we \texttt{smuggle} two
utilities for interacting with local files in the code below. The
\texttt{smuggle} statement in line 8 loads the \texttt{is\_file()}
function from the Python standard library's \texttt{os.path}
module. Since standard library modules are included with all Python
distributions, this line is functionally equivalent to an
\texttt{import} statement and does not need or benefit from an onion
comment. Line 9 loads the \texttt{joblib} package~\cite{Varo10},
installing it first, if necessary. Since \texttt{joblib}'s I/O
interface has historically remained stable and backwards-compatible
across releases, requiring a particular exact version
would likely be unnecessarily restrictive. However, it is possible a
\textit{future} release could introduce some breaking change.  The
onion comment in line 9 helps ensure the analysis notebook continues
to run properly in the future by limiting allowable versions to those
already released when the code was written:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example2}
\end{center}
Line 7 then uses the \texttt{davos.config} object to enable
Davos's \texttt{auto\_rerun} option before smuggling the next
two packages: \texttt{NumPy} and
\texttt{pandas}. Because these packages rely heavily on custom C data
types, loading the particular versions from the onion comments may
require restarting the notebook kernel if different versions had been previously
imported during the same interpreter session (see
Sec.~\ref{subsec:config}).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example3}
\end{center}

>>>>

Across different versions of a given package, various modules, functions,
and other objects may be updated, removed, renamed, or otherwise altered. In
addition to changing the behaviors of active computations, these changes can
render saved objects created using one version of a package incompatible with
other versions of the same package. For example, the popular
\texttt{pandas}~\cite{McKi10} library originally included the \texttt{Panel} data
structure for storing 3-dimensional arrays. In version 0.20.0, however, the
\texttt{Panel} class was deprecated, and in version 0.25.0, it was removed
entirely. Suppose a user had a dataset stored in a \texttt{Panel} object
(created using an older version of \texttt{pandas}) and had saved it to their
disk (e.g., for later reuse or to share with other users) by serializing the
\texttt{Panel} with Python's \texttt{pickle} protocol. The \texttt{pickle}
protocol is a popular built-in method of persisting data in Python that allows
users to save, share, and load arbitrary objects. However, in order to
successfully ``unpickle'' (i.e., load and restore) a ``pickled'' (i.e., previously saved)
object, that object's class must be defined in and importable from the same
module as it was when the object was originally saved. Thus, because of the \texttt{Panel} class's
removal, the user's dataset could not be read by any version of \texttt{pandas}
from 0.25.0 onward. These incompatibilities are also not limited solely to
traditional forms of data. For example, saved model states and other objects
may reference modules, functions, attributes, classes, or other objects that
may not be identical (or even present) across all versions of their associated
packages.

The example provided in Figure~\ref{fig:illustrative-example} demonstrates how
Davos can be used to circumvent these incompatibilities by
temporarily switching between different versions of the same package within a single runtime.
%carefully controlling which versions of each package are used in different parts of the notebook.
The example shows how a dataset and model that require
now-incompatible components of the \texttt{pandas} and
\texttt{scikit-learn}~\cite{PedrEtal11} libraries can be loaded in (using older
versions of each package) and used alongside more recent versions of each
package that provide new and improved functionality. When included at the top
of a Jupyter notebook, the code in Figure~\ref{fig:illustrative-example}
ensures that these objects will be loaded successfully and analyzed using the
same set of package versions no matter when or by whom the notebook is run.

After installing and importing Davos (lines 1--2), we first use the \texttt{davos.require\_python()} function to constrain the Python version used to run the notebook (see Sec.~\ref{subsec:toplevel}).
As described above, the example code in Figure \ref{fig:illustrative-example} loads two different versions of the \texttt{pandas} library: first, an older version needed to access a dataset saved in an outmoded format, then a newer one for use throughout the remainder of the notebook.
We therefore want to make sure upfront (in line 6) that the notebook's Python version falls within the range of versions that both of these two versions of \texttt{pandas} support.
%Line 6 therefore ensures upfront that the notebook's Python version falls within the overlap between the ranges of Python versions that these two versions of \texttt{pandas} support.
If it does not, the function in line 6 will raise an error that includes a message to this effect (lines 4--5).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example1}
\end{center}

Next, in lines 8--9, we \texttt{smuggle} two
utilities for interacting with local files in the code below. The
\texttt{smuggle} statement in line 8 loads the \texttt{is\_file()}
function from the Python standard library's \texttt{os.path}
module. Standard library modules are included with all Python
distributions, so this line is functionally equivalent to an
\texttt{import} statement and does not need or benefit from an onion
comment (since there is no chance it will need to be installed).
Line 9 then loads the \texttt{joblib} package~\cite{Varo10},
installing it into the notebook's project directory, if necessary. Since \texttt{joblib}'s I/O
interface has historically remained stable and backwards-compatible
across releases, requiring a particular exact version
would likely be unnecessarily restrictive. However, it is possible a
\textit{future} release could introduce some breaking change.  The
onion comment in line 9 helps ensure the analysis notebook continues
to run properly in the future by limiting allowable versions to those
already released when the code was written:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example2}
\end{center}
Line 7 then uses the \texttt{davos.config} object to enable
Davos's \texttt{auto\_rerun} option before smuggling the next
two packages: \texttt{NumPy} and
\texttt{pandas}. Because these packages rely heavily on custom C data
types, loading the particular versions from the onion comments may
require restarting the notebook kernel if different versions had been previously
imported during the same interpreter session (see
Sec.~\ref{subsec:config}).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example3}
\end{center}
